---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    css: custom.css
---

```{r setup, include=FALSE}
  install.packages("ggplot2", repos="http://cran.rstudio.com/")
  install.packages("xtable", repos="http://cran.rstudio.com/")
  library(ggplot2)
  library(xtable)
```

## Loading and preprocessing the data
We read in the data from the .zip file distributed with the base repository.

```{r origData, warning=FALSE, message=FALSE}

  ## Read in the data - it comes in our project's git repo as a zip of a .csv
  ## extract it and save a copy of the original data
  origData <- read.csv(unz(paste(getwd(), "/activity.zip", sep = ""), "activity.csv"))
```

## What is mean total number of steps taken per day?
```{r histRawMeanStepsDay, warning=FALSE, message=FALSE}
    ## Create a frame that contains some useful breakdowns
  origView <- data.frame(date = as.Date(levels(origData$date)), 
                      steps = aggregate(x = origData$steps, 
                                        by = list(Date=origData$date), 
                                        FUN = sum, na.rm = F)[,2], 
                      meanSteps = aggregate(x = origData$steps, 
                                        by = list(Date=origData$date), 
                                        FUN = mean, na.rm = F)[,2], 
                      medianSteps = aggregate(x = origData$steps, 
                                        by = list(Date=origData$date), 
                                        FUN = median, na.rm = F)[,2])
  
  ggplot(data = origView, 
         aes(origView$date, origView$steps)) + 
          geom_bar(stat="identity") + 
          labs(x = "Date", y = "Steps") + 
          scale_x_date(breaks = "week") +
          theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  origXtab <- data.frame(Date = as.character(origView$date), 
                      Total = origView$steps, 
                      Average = origView$meanSteps, 
                      Median = origView$medianSteps)
```
```{r tabRawStats, results='asis',  warning=FALSE, message=FALSE}  
  print(xtable(origXtab, align = c("", "c", "c", "c", "c")), 
        type = "html", 
        include.rownames=FALSE)
```

## What is the average daily activity pattern?
A graph of the average steps taken per interval.
```{r lineAvgStepsInterval}
  avgByInterval <- data.frame(Interval = unique(origData$interval),
                              AvgSteps = aggregate(x = origData$steps,
                                                by = list(interval = origData$interval),
                                                FUN = mean, 
                                                na.rm = T)[,2])
  plot(x = avgByInterval$Interval, 
       y = avgByInterval$AvgSteps, 
       type = "l", 
       xlab = "Interval", ylab = "Average Steps")
```

```{r, echo=FALSE}
  averageMaxInterval <- subset(avgByInterval, AvgSteps == max(avgByInterval$AvgSteps))
```
Interval `r averageMaxInterval$Interval` shows the maximum average number of steps with `r averageMaxInterval$AvgSteps`

## Imputing missing values
We impute missing values by using our previously generated interval-averages to demonstrate how 'NA' values (and a strategy
to replace them) can alter our outcomes.  We use a crude method of replacement - using the overall average number of steps
in our dataset, to replace 'NA' / unrecorded values.  This is intentionally 'bad' to demonstrate the impact of a poor
strategy choice.  
```{r histImputedMeanStepsDay, warning=FALSE, message=FALSE}
  overallMean <- mean(origData$steps, na.rm=T)
  imputedData <- data.frame(origData)
  naCount <- length(imputedData$steps[which(is.na(imputedData$steps))])
  imputedData$steps[which(is.na(imputedData$steps))] <- overallMean

   imputedView <- data.frame(date = as.Date(levels(imputedData$date)), 
                      steps = aggregate(x = imputedData$steps, 
                                        by = list(Date=imputedData$date), 
                                        FUN = sum, na.rm = F)[,2], 
                      meanSteps = aggregate(x = imputedData$steps, 
                                        by = list(Date=imputedData$date), 
                                        FUN = mean, na.rm = F)[,2], 
                      medianSteps = aggregate(x = imputedData$steps, 
                                        by = list(Date=imputedData$date), 
                                        FUN = median, na.rm = F)[,2])
  
  ggplot(data = imputedView, 
         aes(imputedView$date, imputedView$steps)) + 
          geom_bar(stat="identity") + 
          labs(x = "Date", y = "Steps") + 
          scale_x_date(breaks = "week") +
          theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  imputedXtab <- data.frame(Date = as.character(imputedView$date), 
                      Total = imputedView$steps, 
                      Average = imputedView$meanSteps, 
                      Median = imputedView$medianSteps)
```
```{r tabImputedStats, results='asis',  warning=FALSE, message=FALSE}  
  print(xtable(imputedXtab, align = c("", "c", "c", "c", "c")), 
        type = "html", 
        include.rownames=FALSE)
```

The above histogram and table are the result of replacing `r naCount` uncollected data points in the original dataset with the 
value `r overallMean`.  A few changes of note:

  * Some medians are now non-zerio
  * There are no days reporting zero total steps
  * While ceilings are not noticeably increased, floors are significantly higher
  
  
## Are there differences in activity patterns between weekdays and weekends?
We use the imputedView calculated above with a bit of factor manipulation to introspect into activity patterns on 
weekend days (Saturday, Sunday) vs week days.
```{r lineImputedByInterval}
  ## Alter the existing 'date' column into the 'weekdays()' format
  # Levels: Friday Monday Saturday Sunday Thursday Tuesday Wednesday
  imputedData$date <- factor(weekdays(as.Date(as.character(imputedData$date))))
  
  ## Alter the (now factored) date field again to read either 'Weekday' or 'Weekend'
  dayTypeFactor <- c("Weekday", # Friday 
                     "Weekday", # Monday
                     "Weekend", # Saturday 
                     "Weekend", # Sunday 
                     "Weekday", # Thursday 
                     "Weekday", # Tuesday 
                     "Weekday") # Wednesday
  
  ## Compress them
  imputedData$date <- factor(dayTypeFactor[imputedData$date])
  
  ## Aggregate by factors over intervals
  avgImputedByInterval <- data.frame(
                            aggregate(
                              x = imputedData$steps,
                              by = list(
                                interval = imputedData$interval, 
                                date = imputedData$date),
                              FUN = mean, 
                              na.rm = T))
  
  ## Create a facet grid based on our new factors ('Weekday' and 'Weekend')
  ggplot(data = avgImputedByInterval, 
         aes(avgImputedByInterval$interval, avgImputedByInterval$x)) + 
          geom_line() + 
          labs(x = "Interval", y = "Average Steps Taken") +
          facet_wrap(~ date, ncol = 1)
```
  
As the graph illustrates, weekdays show more activity earlier and later in the day (perhaps before and after work / school / etc)
with weekends showing more consistent activity in the center of the day (perhaps during leisure activities).